Interpretable Machine Learning for Science \\ with PySR and SymbolicRegression.jl

### 2023-05-02 src/tex/ms.tex
Grammar fixes

\twocolumn[
\begin{@twocolumnfalse}
\maketitle
\begin{abstract}
\pysr is an open-source library for practical \textit{symbolic regression}, a type of machine learning which aims to discover human-interpretable symbolic models.
\pysr was developed to democratize and popularize symbolic regression for the sciences, and is built on a high-performance distributed backend, a flexible search algorithm, and interfaces with several deep learning packages.
\pysr's internal search algorithm is a multi-population evolutionary algorithm, which consists of a unique evolve-simplify-optimize loop, designed for optimization of unknown scalar constants in newly-discovered empirical expressions.
\pysr's backend is an extremely optimized Julia library \pysrjl, which can be used directly from Julia. It is capable of fusing user-defined operators into SIMD kernels at runtime, performing automatic differentiation, and distributing populations of expressions to thousands of cores across a cluster.
In describing this software, \we also introduce a new benchmark, ``\textit{EmpiricalBench},’’ to quantify the applicability of symbolic regression algorithms in science.
This benchmark measures recovery of historical empirical equations from original and synthetic datasets.
\end{abstract}
\end{@twocolumnfalse}
]

@@

This brings us to an optimization problem known ``symbolic regression,'' or SR.
In this paper, \we describe an algorithm for performing \sr, and describe software dedicated to equation discovery in the sciences.

@@

\paragraph{Equation Discovery for Science.} 
Today \eureqa is a closed-source proprietary tool, and has fallen out of common use in the sciences.
In addition to other early work such as~\cite{debFastElitistNondominated2000,debFastElitistMultiobjective2002,davidsonSymbolicNumericalRegression2003,cranmerPhysicsGPGeneticProgramming2005},
there have recently been a significant number of exciting new developments in \sr as well as \sr for science.
These have largely been focused on genetic algorithms on expressions represented as trees, and include core advances to genetic algorithms, such as~\cite{lacavaEpsilonLexicaseSelectionRegression2016,lacavaProbabilisticMultiobjectiveAnalysis2018}, which propose modified strategies for selecting an individual to mutate and~\cite{virgolinScalableGeneticProgramming2017,virgolinImprovingModelbasedGenetic2021} which use an adaptive form of recombination based on emerging statistical patterns.
New strategies for \sr altogether have also been developed:
in particular, \our previous work~\cite{cranmerLearningSymbolicPhysics2019,cranmerDiscoveringSymbolicModels2020,cranmerDisentangledSparsityNetworks2021} proposes \sr as a way of interpreting a neural network.
Neural networks can efficiently find low-dimensional patterns in high-dimensional data, and so this mixed strategy also presents a way of performing \sr on high-dimensional or non-tabular data.

@@

However, here \we argue that many existing algorithms lack certain features which make them practical in a broad setting for discovering interpretable equations in science.
The requirements for ``scientific \sr'' are significantly more difficult than for \sr applied to synthetic datasets, which many algorithms are implicitly trained on via synthetic benchmarks.
We list these below.

@@

\paragraph{Contribution.}
In this paper, \we introduce \pysr, a Python and Julia package for Scientific Symbolic Regression. The Python package is available under \texttt{pysr} on PyPI and Conda\footnote{\myhref{https://github.com/MilesCranmer/PySR}{github.com/MilesCranmer/PySR}} and the Julia library under \texttt{SymbolicRegression}\footnote{\myhref{https://github.com/MilesCranmer/SymbolicRegression.jl}{github.com/MilesCranmer/SymbolicRegression.jl}} on the Julia package registry.
Every part of \pysr is written with scientific discovery in mind, as the entire point of creating this package was to enable the authors of this paper to use this tool in their own field for equation discovery.
All of the requirements of discovering an empirical equation for science, as listed above, are satisfied by \pysr.
\We discuss this in detail in the following sections.

@@

First, \we give an in-depth discussion of the internal algorithm in \pysr in \cref{sec:algorithm}, and then describe the software implementation in \cref{sec:software}.
We then discuss a few additional implementation features in \cref{sec:interfaces}, \cref{sec:custom_operators}, and \cref{sec:custom_losses}.

@@

This algorithm is computationally efficient and allows for massive parallelization by splitting step~\ref{step:sampling} into groups of individuals which can evolve independently, with asynchronous ``migration'' between the groups.
For the first of the modifications, \we replace the eldest member of a population in step~\ref{step:mutate}, rather than the individual with the lowest fitness, which is known as ``age-regularized'' evolution~\cite{realRegularizedEvolutionImage2019}.
We record the age from the Unix time at which the individual was created during step~\ref{step:mutate}.
Its inclusion in \pysr was motivated by the impressive Neural Architecture Search results of~\cite{realRegularizedEvolutionImage2019,realAutoMLZeroEvolvingMachine2020a} --- though it is worth noting similar regularizations used earlier in SR, e.g.,~\cite{hornbyALPSAgelayeredPopulation2006} and~\cite{schmidtAgefitnessParetoOptimization2010} --- which found this simple modification could prevent early convergence without hurting performance.
This prevents the population from specializing too quickly and getting stuck in a local minimum of the search space.
The implementation in~\cite{realAutoMLZeroEvolvingMachine2020a} uses $p=1$ and explores $4\leq n_s \leq 64$, though here \we allow $p$ to be less than $1$, among several other differences.

\paragraph{Modifications.}
\Our algorithm in \pysr, first released publicly in 2020~\cite{cranmerPySRFastParallelized2020a}, makes three changes to this evolutionary algorithm.
\We apply simulated annealing~\cite{kirkpatrickOptimizationSimulatedAnnealing1983} to step~\ref{step:select}: given a temperature $T\in[0, 1]$, a mutation is rejected with some probability related to the change in fitness after mutation.
The probability for rejection is $p=\exp(\frac{L_F - L_E}{\alpha T})$, for $L_F$ and $L_E$ the fitness of the mutated and original individual respectively, and $\alpha$ a hyperparameter.
This allows the evolution to alternate between \textit{high temperature} and \textit{low temperature} phases, with the high temperature phases increasing diversity of individuals and the low temperature phases narrowing in on the fittest individuals.
$\alpha$ controls the scale of temperature, with $\alpha\rightarrow\infty$ being equivalent to regular tournament selection, and $\alpha\rightarrow 0$ rejecting all mutations which lower fitness.
Since the value of $p$ also controls the diversity of the population, \we can also couple the value of $p$ to temperature, setting a larger value of $p$ for a higher temperature, and a smaller value of $p$ for a lower temperature.
Simulated annealing is a simple yet powerful strategy for optimizing over discrete spaces, and \we find in \cref{sec:evaluation} that simulated annealing can significantly speed up the search process.

The second modification \we make to tournament selection is specific to the individuals \we wish to consider: mathematical expressions.
\We embed the genetic algorithm inside an \textit{evolve-simplify-optimize loop}.
``Evolve'' is a repeated application of tournament selection-based evolution for a set number of mutations.
In other words, this is an entire evolution of a population.
``Simplify'' here refers to equation simplification---equations are simplified to an equivalent form during the loop using a set of algebraic equivalencies (since this happens infrequently enough, it tends to not harm discovery by constraining the search space).
The ``Optimize'' part of this algorithm is a few iterations of a classical optimization algorithm (BFGS~\cite{broydenConvergenceClassDoublerank1970} is used by default, though any optimizer in the \texttt{Optim.jl} package can be used~\cite{mogensenOptimMathematicalOptimization2018}) to optimize constants in every equation explicitly.
This part of the algorithm significantly improves the discovery of equations containing real constants, which is essential for practical application of \sr to science.
It has been shown since at least~\cite{topchyFasterGeneticProgramming2001} that performing local gradient searches on numerical constants in expressions can vastly improve the performance of \sr---\we closely integrate an optimized and stochastic version of this.

@@

Finally, the third modification \we make is the use of a novel \textit{adaptive parsimony} metric.
But first, \we give a definition of complexity.

@@

    However, complexity in \pysr is a completely user-configurable. \We argue that the best definition for a ``simple'' expression is the one that is most interpretable by the user. An expression containing a dilogarithm may look commonplace to the particle physicist---and in fact be a great prior over the space of models for certain particle physics data---but extremely unusual to the fluid dynamicist, and perhaps not coincidentally a bad choice of prior for fluids data.
}

\noindent The traditional mechanism for penalizing complexity would be to use a constant \textit{parsimony}, which would involve adding a loss term equal to the complexity of an expression times a constant factor:
\begin{equation}
    \ell(E) = \ell_\text{pred}(E) + (\text{parsimony}) \cdot C(E),
\end{equation}
for $C(E)$ the complexity of an expression $E$ and $\ell_\text{pred}$ the predictive loss.
Instead, \we adaptively tune a per-complexity penalty such that the number of expressions at each complexity is approximately the same.
This is expressed roughly as:
\begin{equation}
    \ell(E) = \ell_\text{pred}(E) \cdot \exp( \text{frecency}[C(E)]),
\end{equation}
where the ``frecency'' of $C(E)$ is some combined measure of the frequency and recency of expressions occurring at complexity $C(E)$ in the population.
In \pysr, we simply count the number of expressions with a moving window in time, and divide by a tunable constant.
Thus, this is a simple way to punish complexities adaptively by how frequent they are in the population.
This encourages the evolutionary search to explore the problem from simple, less-accurate expressions, as well as from complex, more accurate expressions.
This qualitatively seems to help alleviate situations where the search specializes too early in the wrong functional form, and can only seem to make small iterations to it, and is discouraged from starting from scratch.

Pseudocode for the outer loop loop of \pysr is given in \cref{alg:pysr}, which includes the migration steps shown in \cref{fig:migration}.
\Cref{alg:evolve} gives pseudocode for the main \textit{evolve-simplify-optimize loop}, formalizing parts of the cartoon in \cref{fig:evolution_loop}.
\Cref{alg:mutate} outlines the set of mutations, supplemented with simulated annealing.
Finally, \cref{alg:tournament} describes the tournament selection strategy.

@@

## Second, \pysr allows one to specify a set of weights for each input datapoint. This could be used, for instance, if a user knows the uncertainty in a measurement beforehand to be $\sigma>0$, for $\sigma$ the standard deviation of the measurement, and thus weights each datapoint in the loss by the signal-to-noise $1/\sigma^2$.

## Thirdly, the user can define a custom likelihood to optimize, which can also take into account weights. This is explained in the following point.
# \textit{Custom losses.}
\pysr supports arbitrary user-defined loss functions, passed as either a string or a function object.
These take a scalar prediction and target, and optionally a scalar weight, and should return a real number greater than 0.
These per-point losses are then summed over the dataset.
With this, the user is capable of defining arbitrary regression objectives, custom likelihoods (defined as log-likelihood of a single datapoint), and classification-based losses.
This also allows one to define implicit objectives.
# \textit{Custom operators.}
One of the most powerful features of \pysr is the ability to define custom operators.
Different domains of science have functions unique to their field which appear in many formulae and hold a specific meaning.
It is therefore very important that these operators be available in an \sr library, and so \pysr allows custom user-defined unary or binary functions.
So long as a function can be defined as either
$f:\mathbb{R}\rightarrow \mathbb{R}$ or $\mathbb{R}^2\rightarrow\mathbb{R}$, it can be used as a user-defined operator.
\pysr \ul{does not treat built-in operators any differently than user-defined ones.}
Apart from simplification strategies, \pysr does not know the difference between the $+$ operator and a Bessel function.
# \textit{Feature selection.}
Similar to the Gaussian process preprocessing step for denoising, \pysr also uses a simple dimensionality reduction preprocessing step.
Given a user-defined number of features to select, \pysr uses a gradient-boosting tree algorithm to first fit the dataset, then select the most important features.
These features are fed into the main search loop.
# \textit{Constraints.}
\pysr allows various hard constraints to be specified for discovered expressions.
These are enforced at every mutation: if a constraint is violated, the mutation is rejected. A few are available with the API:
## The maximum size of an expression (the number of instances of operators, variables, and constants), which gives an overall bound on the search space.
## The maximum depth of an expression, which can be used to control how deeply nested the resultant expression is.
## The maximum size of a subexpression in a specific operator.
This specifies the maximum size of an expression within an operator's arguments.
For example, specifying that the $\wedge$ operator has maximum argument size of $(-1, 3)$ means that its base can have any size expression (indicated by -1), while the expression in its exponent can only have max size of up to 3.
This operator-specific constraint can drastically reduce the complexity of discovered expressions.
## The maximum number of nests of a particular operator combination. For example, the nested constraint \texttt{\{sin: \{sin: 0, cos: 1\}\}} would indicate that $\sin(\cos(x))$ is allowed, but $\sin(\cos(\cos(x)))$ and $\sin(\sin(x))$ are not.
# \textit{Additional data structures}. See deep learning interface \cref{sec:interfaces}.
# \textit{Integrals and dynamical systems}. See deep learning interface \cref{sec:interfaces}.
\end{easylist}

@@

\pysr is parallelized by populations of expressions which evolve independently, which was described as the basis of allowing large-scale parallelization in~\cite{goldbergComparativeAnalysisSelection1991}.
In this framework, a population of expressions is dispatched to a worker (either a thread, if single-node, or a process, if multi-node), where it then evolves over many iterations.
Following a batch of iterations, this population is returned to the head worker, which performs several tasks:
(1) record the best expressions, at each complexity, found in the population as part of a global ``hall of fame;''
(2) record the current expressions so they can ``migrate'' to other populations; and, finally,
(3) randomly ``migrate'' expressions from other populations, and the global hall of fame, to this population.
After these steps are finished, this population is once again dispatched to a worker where it undergoes further evolution.
This parallelization is asynchronous, and different populations may complete evolution at different speeds without slowing down others.
The technique is largely similar to the method described in~\cite{schmidtDistillingFreeFormNatural2009}, although here \we have modified the migration step to also migrate based on a global, permanent hall of fame, rather than exclusively migrating current population members.
Reintroducing hall of fame members can significantly speed up the search.

@@

A similar technique---evaluating the string representation for an expression---is used for exporting expressions to a callable format in NumPy, PyTorch, and JAX~\cite{numpy,torch,jax}---each of which take vector input.
For NumPy, constants in an expression are embedded inside the callable function.
However, for PyTorch and JAX, since one may wish to re-optimize the constants in an expression (for example, as is done in~\cite{lemosRediscoveringNewtonGravity2022}, the constants are trainable.
In PyTorch, this amounts to creating a PyTorch \texttt{nn.Parameter} for the constants, which causes PyTorch to track gradients with respect to the parameters, so they can be trained via gradient descent.
Since JAX is a functional language, the parameters are instead exported as a vector, which the user then will pass into the callable function.
This is similar to how deep learning frameworks in JAX are trained.

@@

In \cref{sec:intro_pysr}, \we discussed practical issues with discovering symbolic expressions in the sciences.
Here, \we present a comparison of \pysr with existing tools in terms of addressing these concerns.

@@

First, \we present a high level comparison of the various \sr tools available.
These qualitative points are made using the desirable features given in \cref{sec:intro_pysr}.

@@

Existing \sr benchmarks have certain simplifications compared to the datasets used to historically discover well-known empirical equations.
When discovering a new expression, one does not actually know the physical constants in the expression, and one must have to learn real constants.
For example, in the Feynman benchmark dataset of~\cite{udrescuAIFeynmanPhysicsinspired2020}, all expressions are listed with the associated physical constants---but these constants originally had to be discovered along with the equation.
Similarly, in the commonly-used synthetic ``Nguyen'' benchmark~\cite{nguyenquanguySemanticallybasedCrossoverGenetic2011}, which includes expressions such as $F_9=\sin(x)+\sin(y^2)$, there are no non-integral constants.
The ``SRBench'' competition of~\cite{defrancaInterpretableSymbolicRegression2023} was meticulous in its comparisons of different methods, but only contained a single real-world task (with an unknown ground truth), with all other expressions being synthetic.
This benchmark improved on others in that it included tunable amounts of noise.
However, the noise models are very simple: perfectly Gaussians and including no heteroscedasticity.
Although such synthetic benchmarks have their uses in \sr research, these types of benchmarks are often not entirely indicative of the challenges faced in real-world scientific discovery.
Therefore, here \we introduce a new benchmark which attempts to accurately portray the empirical equation discovery step of science.
Every equation in this dataset is a real empirical expression that a scientist has at one point discovered from experimental, noisy, and imperfect data.

@@

\We fork the SRBench competition repository~\cite{defrancaInterpretableSymbolicRegression2023}, as the authors have managed to perform the impressive task of aggregating a large variety of existing methods into a single repository with a common API.
However, here we aim to study the entire Pareto front of each algorithm, rather than only a single expression, as was done in SRBench.
By reviewing the discovered expressions over the entire Pareto front, we are not sensitive to the somewhat arbitrary choice of selection criteria, which differs by method. 
We can more accurately gauge performance of the algorithms themselves.
Thus, we modify the interface of every algorithm to return a Pareto front of expressions, or, in the cases where it was unavailable, instead we return a list of several candidate expressions.
\We were able to successfully test the algorithms \pysr, Operon~\cite{burlacuOperonEfficientGenetic2020}, DSR~\cite{petersenDeepSymbolicRegression2021}, EQL~\cite{sahooLearningEquationsExtrapolation2018}, QLattice~\cite{brolosApproachSymbolicRegression2021}, and SR-Transformer~\cite{kamiennyEndtoendSymbolicRegression2022} on \bench.
Other codes were either not included in SRBench already, incompatible with our tests, or were otherwise unable to be configured on \our system after significant effort.

@@

\We emphasize that the stability of an \sr algorithm and software implementation itself has central importance to practical use in science, and therefore, unstable runs which either produce an undefined expression, or crash, are \textit{included} in the results dataset and the primary score.
However, one should also note that because \we are the authors of \pysr, we are therefore more likely to be running in a stable environment than, say, a package which \we are unfamiliar with.
For example, as seen in \cref{tab:results}, DSR sees improved results if one ignores the failed runs, although there were still some expressions it struggled on.
Thus, detailed itemizations of these results are given in \cref{tab:results}, to allow for a more nuanced understanding of the results.
Also note that this benchmark does not explicitly measure other metrics which are considered in other benchmarks, such as number of evaluations (where, e.g., genetic algorithms are very inefficient), nor does it consider the numerical accuracy of the expressions.

>>>>

\twocolumn[\begin{@twocolumnfalse}
\maketitle
\begin{abstract}
\pysr is an open-source library for practical \textit{symbolic regression}, a type of machine learning which aims to discover human-interpretable symbolic models.
\pysr was developed to democratize and popularize symbolic regression for the sciences, and is built on a high-performance distributed backend, a flexible search algorithm, and interfaces with several deep learning packages.
\pysr's internal search algorithm is a multi-population evolutionary algorithm, which consists of a unique evolve-simplify-optimize loop, designed for optimization of unknown scalar constants in newly-discovered empirical expressions.
\pysr's backend is an extremely optimized Julia library \pysrjl, which can be used directly from Julia. It is capable of fusing user-defined operators into SIMD kernels at runtime, performing automatic differentiation, and distributing populations of expressions to thousands of cores across a cluster.
In describing this software, we also introduce a new benchmark, ``\textit{EmpiricalBench},’’ to quantify the applicability of symbolic regression algorithms in science.
This benchmark measures recovery of historical empirical equations from original and synthetic datasets.
\end{abstract}
\end{@twocolumnfalse}
]

@@

This brings us to an optimization problem known ``symbolic regression,'' or SR.
In this paper, we describe an algorithm for performing \sr, and describe software dedicated to equation discovery in the sciences.

@@

\paragraph{Equation Discovery for Science.} 
Today \eureqa is a closed-source proprietary tool, and has fallen out of common use in the sciences.
In addition to other early work such as~\cite{debFastElitistNondominated2000,debFastElitistMultiobjective2002,davidsonSymbolicNumericalRegression2003,cranmerPhysicsGPGeneticProgramming2005},
there have recently been a significant number of exciting new developments in \sr as well as \sr for science.
These have largely been focused on genetic algorithms on expressions represented as trees, and include core advances to genetic algorithms, such as~\cite{lacavaEpsilonLexicaseSelectionRegression2016,lacavaProbabilisticMultiobjectiveAnalysis2018}, which propose modified strategies for selecting an individual to mutate and~\cite{virgolinScalableGeneticProgramming2017,virgolinImprovingModelbasedGenetic2021} which use an adaptive form of recombination based on emerging statistical patterns.
New strategies for \sr altogether have also been developed:
in particular, our previous work~\cite{cranmerLearningSymbolicPhysics2019,cranmerDiscoveringSymbolicModels2020,cranmerDisentangledSparsityNetworks2021} proposes \sr as a way of interpreting a neural network.
Neural networks can efficiently find low-dimensional patterns in high-dimensional data, and so this mixed strategy also presents a way of performing \sr on high-dimensional or non-tabular data.

@@

However, here we argue that many existing algorithms lack certain features which make them practical in a broad setting for discovering interpretable equations in science.
The requirements for ``scientific \sr'' are significantly more difficult than for \sr applied to synthetic datasets, which many algorithms are implicitly trained on via synthetic benchmarks.
We list these below.

@@

\paragraph{Contribution.}
In this paper, we introduce \pysr, a Python and Julia package for Scientific Symbolic Regression. The Python package is available under \texttt{pysr} on PyPI and Conda\footnote{\myhref{https://github.com/MilesCranmer/PySR}{github.com/MilesCranmer/PySR}} and the Julia library under \texttt{SymbolicRegression}\footnote{\myhref{https://github.com/MilesCranmer/SymbolicRegression.jl}{github.com/MilesCranmer/SymbolicRegression.jl}} on the Julia package registry.
Every part of \pysr is written with scientific discovery in mind, as the entire point of creating this package was to enable the authors of this paper to use this tool in their own field for equation discovery.
All of the requirements of discovering an empirical equation for science, as listed above, are satisfied by \pysr.
We discuss this in detail in the following sections.

@@

First, we give an in-depth discussion of the internal algorithm in \pysr in \cref{sec:algorithm}, and then describe the software implementation in \cref{sec:software}.
We then discuss a few additional implementation features in \cref{sec:interfaces}, \cref{sec:custom_operators}, and \cref{sec:custom_losses}.

@@

This algorithm is computationally efficient and allows for massive parallelization by splitting step~\ref{step:sampling} into groups of individuals which can evolve independently, with asynchronous ``migration'' between the groups.
For the first of the modifications, we replace the eldest member of a population in step~\ref{step:mutate}, rather than the individual with the lowest fitness, which is known as ``age-regularized'' evolution~\cite{realRegularizedEvolutionImage2019}.
We record the age from the Unix time at which the individual was created during step~\ref{step:mutate}.
Its inclusion in \pysr was motivated by the impressive Neural Architecture Search results of~\cite{realRegularizedEvolutionImage2019,realAutoMLZeroEvolvingMachine2020a} --- though it is worth noting similar regularizations used earlier in SR, e.g.,~\cite{hornbyALPSAgelayeredPopulation2006} and~\cite{schmidtAgefitnessParetoOptimization2010} --- which found this simple modification could prevent early convergence without hurting performance.
This prevents the population from specializing too quickly and getting stuck in a local minimum of the search space.
The implementation in~\cite{realAutoMLZeroEvolvingMachine2020a} uses $p=1$ and explores $4\leq n_s \leq 64$, though here we allow $p$ to be less than $1$, among several other differences.

\paragraph{Modifications.}
Our algorithm in \pysr, first released publicly in 2020~\cite{cranmerPySRFastParallelized2020a}, makes three changes to this evolutionary algorithm.
We apply simulated annealing~\cite{kirkpatrickOptimizationSimulatedAnnealing1983} to step~\ref{step:select}: given a temperature $T\in[0, 1]$, a mutation is rejected with some probability related to the change in fitness after mutation.
The probability for rejection is $p=\exp(\frac{L_F - L_E}{\alpha T})$, for $L_F$ and $L_E$ the fitness of the mutated and original individual respectively, and $\alpha$ a hyperparameter.
This allows the evolution to alternate between \textit{high temperature} and \textit{low temperature} phases, with the high temperature phases increasing diversity of individuals and the low temperature phases narrowing in on the fittest individuals.
$\alpha$ controls the scale of temperature, with $\alpha\rightarrow\infty$ being equivalent to regular tournament selection, and $\alpha\rightarrow 0$ rejecting all mutations which lower fitness.
Since the value of $p$ also controls the diversity of the population, we can also couple the value of $p$ to temperature, setting a larger value of $p$ for a higher temperature, and a smaller value of $p$ for a lower temperature.
Simulated annealing is a simple yet powerful strategy for optimizing over discrete spaces, and we find in \cref{sec:evaluation} that simulated annealing can significantly speed up the search process.

The second modification we make to tournament selection is specific to the individuals we wish to consider: mathematical expressions.
We embed the genetic algorithm inside an \textit{evolve-simplify-optimize loop}.
``Evolve'' is a repeated application of tournament selection-based evolution for a set number of mutations.
In other words, this is an entire evolution of a population.
``Simplify'' here refers to equation simplification---equations are simplified to an equivalent form during the loop using a set of algebraic equivalencies (since this happens infrequently enough, it tends to not harm discovery by constraining the search space).
The ``Optimize'' part of this algorithm is a few iterations of a classical optimization algorithm (BFGS~\cite{broydenConvergenceClassDoublerank1970} is used by default, though any optimizer in the \texttt{Optim.jl} package can be used~\cite{mogensenOptimMathematicalOptimization2018}) to optimize constants in every equation explicitly.
This part of the algorithm significantly improves the discovery of equations containing real constants, which is essential for practical application of \sr to science.
It has been shown since at least~\cite{topchyFasterGeneticProgramming2001} that performing local gradient searches on numerical constants in expressions can vastly improve the performance of \sr---we closely integrate an optimized and stochastic version of this.

@@

Finally, the third modification we make is the use of a novel \textit{adaptive parsimony} metric.
But first, we give a definition of complexity.

@@

    However, complexity in \pysr is a completely user-configurable. We argue that the best definition for a ``simple'' expression is the one that is most interpretable by the user. An expression containing a dilogarithm may look commonplace to the particle physicist---and in fact be a great prior over the space of models for certain particle physics data---but extremely unusual to the fluid dynamicist, and perhaps not coincidentally a bad choice of prior for fluids data.
}

\noindent The traditional mechanism for penalizing complexity would be to use a constant \textit{parsimony}, which would involve adding a loss term equal to the complexity of an expression times a constant factor:
\begin{equation}
    \ell(E) = \ell_\text{pred}(E) + (\text{parsimony}) \cdot C(E),
\end{equation}
for $C(E)$ the complexity of an expression $E$ and $\ell_\text{pred}$ the predictive loss.
Instead, we adaptively tune a per-complexity penalty such that the number of expressions at each complexity is approximately the same.
This is expressed roughly as:
\begin{equation}
    \ell(E) = \ell_\text{pred}(E) \cdot \exp( \text{frecency}[C(E)]),
\end{equation}
where the ``frecency'' of $C(E)$ is some combined measure of the frequency and recency of expressions occurring at complexity $C(E)$ in the population.
In \pysr, we simply count the number of expressions with a moving window in time, and divide by a tunable constant.
Thus, this is a simple way to punish complexities adaptively by how frequent they are in the population.
This encourages the evolutionary search to explore the problem from simple, less-accurate expressions, as well as from complex, more accurate expressions.
This qualitatively seems to help alleviate situations where the search specializes too early in the wrong functional form, and can only seem to make small iterations to it, and is discouraged from starting from scratch.

Pseudocode for the outer loop of \pysr is given in \cref{alg:pysr}, which includes the migration steps shown in \cref{fig:migration}.
\Cref{alg:evolve} gives pseudocode for the main \textit{evolve-simplify-optimize loop}, formalizing parts of the cartoon in \cref{fig:evolution_loop}.
\Cref{alg:mutate} outlines the set of mutations, supplemented with simulated annealing.
Finally, \cref{alg:tournament} describes the tournament selection strategy.

@@

## Second, \pysr allows one to specify a set of weights for each input data point. This could be used, for instance, if a user knows the uncertainty in a measurement beforehand to be $\sigma>0$, for $\sigma$ the standard deviation of the measurement, and thus weights each data point in the loss by the signal-to-noise $1/\sigma^2$.

## Thirdly, the user can define a custom likelihood to optimize, which can also take into account weights. This is explained in the following point.
# \textit{Custom losses.}
\pysr supports arbitrary user-defined loss functions, passed as either a string or a function object.
These take a scalar prediction and target, and optionally a scalar weight, and should return a real number greater than 0.
These per-point losses are then summed over the dataset.
With this, the user is capable of defining arbitrary regression objectives, custom likelihoods (defined as log-likelihood of a single data point), and classification-based losses.
This also allows one to define implicit objectives.
# \textit{Custom operators.}
One of the most powerful features of \pysr is the ability to define custom operators.
Different domains of science have functions unique to their field which appear in many formulae and hold a specific meaning.
It is therefore very important that these operators be available in an \sr library, and so \pysr allows custom user-defined unary or binary functions.
So long as a function can be defined as either
$f:\mathbb{R}\rightarrow \mathbb{R}$ or $\mathbb{R}^2\rightarrow\mathbb{R}$, it can be used as a user-defined operator.
\pysr \ul{does not treat built-in operators any differently than user-defined ones.}
Apart from simplification strategies, \pysr does not know the difference between the $+$ operator and a Bessel function.
# \textit{Feature selection.}
Similar to the Gaussian process preprocessing step for denoising, \pysr also uses a simple dimensionality reduction preprocessing step.
Given a user-defined number of features to select, \pysr uses a gradient-boosting tree algorithm to first fit the dataset, then select the most important features.
These features are fed into the main search loop.
# \textit{Constraints.}
\pysr allows various hard constraints to be specified for discovered expressions.
These are enforced at every mutation: if a constraint is violated, the mutation is rejected. A few are available with the API:
## The maximum size of an expression (the number of instances of operators, variables, and constants), which gives an overall bound on the search space.
## The maximum depth of an expression, which can be used to control how deeply nested the resultant expression is.
## The maximum size of a subexpression in a specific operator.
This specifies the maximum size of an expression within an operator's arguments.
For example, specifying that the $\wedge$ operator has maximum argument size of $(-1, 3)$ means that its base can have any size expression (indicated by -1), while the expression in its exponent can only have max size of up to 3.
This operator-specific constraint can drastically reduce the complexity of discovered expressions.
## The maximum number of nests of a particular operator combination. For example, the nested constraint \texttt{\{sin: \{sin: 0, cos: 1\}\}} would indicate that $\sin(\cos(x))$ is allowed, but $\sin(\cos(\cos(x)))$ and $\sin(\sin(x))$ are not.
# \textit{Additional data structures}. See deep learning interface \cref{sec:interfaces}.
# \textit{Integrals and dynamical systems}. See deep learning interface \cref{sec:interfaces}.
\end{easylist}

@@

\pysr is parallelized by populations of expressions which evolve independently, which was described as the basis of allowing large-scale parallelization in~\cite{goldbergComparativeAnalysisSelection1991}.
In this framework, a population of expressions is dispatched to a worker (either a thread, if single-node, or a process, if multi-node), where it then evolves over many iterations.
Following a batch of iterations, this population is returned to the head worker, which performs several tasks:
(1) record the best expressions, at each complexity, found in the population as part of a global ``hall of fame;''
(2) record the current expressions so they can ``migrate'' to other populations; and, finally,
(3) randomly ``migrate'' expressions from other populations, and the global hall of fame, to this population.
After these steps are finished, this population is once again dispatched to a worker where it undergoes further evolution.
This parallelization is asynchronous, and different populations may complete evolution at different speeds without slowing down others.
The technique is largely similar to the method described in~\cite{schmidtDistillingFreeFormNatural2009}, although here we have modified the migration step to also migrate based on a global, permanent hall of fame, rather than exclusively migrating current population members.
Reintroducing hall of fame members can significantly speed up the search.

@@

A similar technique---evaluating the string representation for an expression---is used for exporting expressions to a callable format in NumPy, PyTorch, and JAX~\cite{numpy,torch,jax}---each of which take vector input.
For NumPy, constants in an expression are embedded inside the callable function.
However, for PyTorch and JAX, since one may wish to re-optimize the constants in an expression (for example, as is done in~\cite{lemosRediscoveringNewtonGravity2022}), the constants are trainable.
In PyTorch, this amounts to creating a PyTorch \texttt{nn.Parameter} for the constants, which causes PyTorch to track gradients with respect to the parameters, so they can be trained via gradient descent.
Since JAX is a functional language, the parameters are instead exported as a vector, which the user then will pass into the callable function.
This is similar to how deep learning frameworks in JAX are trained.

@@

In \cref{sec:intro_pysr}, we discussed practical issues with discovering symbolic expressions in the sciences.
Here, we present a comparison of \pysr with existing tools in terms of addressing these concerns.

@@

First, we present a high level comparison of the various \sr tools available.
These qualitative points are made using the desirable features given in \cref{sec:intro_pysr}.

@@

Existing \sr benchmarks have certain simplifications compared to the datasets used to historically discover well-known empirical equations.
When discovering a new expression, one does not actually know the physical constants in the expression, and one must have to learn real constants.
For example, in the Feynman benchmark dataset of~\cite{udrescuAIFeynmanPhysicsinspired2020}, all expressions are listed with the associated physical constants---but these constants originally had to be discovered along with the equation.
Similarly, in the commonly-used synthetic ``Nguyen'' benchmark~\cite{nguyenquanguySemanticallybasedCrossoverGenetic2011}, which includes expressions such as $F_9=\sin(x)+\sin(y^2)$, there are no non-integral constants.
The ``SRBench'' competition of~\cite{defrancaInterpretableSymbolicRegression2023} was meticulous in its comparisons of different methods, but only contained a single real-world task (with an unknown ground truth), with all other expressions being synthetic.
This benchmark improved on others in that it included tunable amounts of noise.
However, the noise models are very simple: perfectly Gaussian and including no heteroscedasticity.
Although such synthetic benchmarks have their uses in \sr research, these types of benchmarks are often not entirely indicative of the challenges faced in real-world scientific discovery.
Therefore, here we introduce a new benchmark which attempts to accurately portray the empirical equation discovery step of science.
Every equation in this dataset is a real empirical expression that a scientist has at one point discovered from experimental, noisy, and imperfect data.

@@

We fork the SRBench competition repository~\cite{defrancaInterpretableSymbolicRegression2023}, as the authors have managed to perform the impressive task of aggregating a large variety of existing methods into a single repository with a common API.
However, here we aim to study the entire Pareto front of each algorithm, rather than only a single expression, as was done in SRBench.
By reviewing the discovered expressions over the entire Pareto front, we are not sensitive to the somewhat arbitrary choice of selection criteria, which differs by method. 
We can more accurately gauge performance of the algorithms themselves.
Thus, we modify the interface of every algorithm to return a Pareto front of expressions, or, in the cases where it was unavailable, instead we return a list of several candidate expressions.
We were able to successfully test the algorithms \pysr, Operon~\cite{burlacuOperonEfficientGenetic2020}, DSR~\cite{petersenDeepSymbolicRegression2021}, EQL~\cite{sahooLearningEquationsExtrapolation2018}, QLattice~\cite{brolosApproachSymbolicRegression2021}, and SR-Transformer~\cite{kamiennyEndtoendSymbolicRegression2022} on \bench.
Other codes were either not included in SRBench already, incompatible with our tests, or were otherwise unable to be configured on our system after significant effort.

@@

We emphasize that the stability of an \sr algorithm and software implementation itself has central importance to practical use in science, and therefore, unstable runs which either produce an undefined expression, or crash, are \textit{included} in the results dataset and the primary score.
However, one should also note that because we are the authors of \pysr, we are therefore more likely to be running in a stable environment than, say, a package which we are unfamiliar with.
For example, as seen in \cref{tab:results}, DSR sees improved results if one ignores the failed runs, although there were still some expressions it struggled on.
Thus, detailed itemization of these results is given in \cref{tab:results}, to allow for a more nuanced understanding of the results.
Also note that this benchmark does not explicitly measure other metrics which are considered in other benchmarks, such as number of evaluations (where, e.g., genetic algorithms are very inefficient), nor does it consider the numerical accuracy of the expressions.
